---
title: "PPR Network Visualization"
author: "Shackett"
date: "`r format(Sys.time(), '%d %B, %Y')`"
format:
  html:
    embed-resources: true
    code-fold: true
    code-summary: "Show the code"
    toc: true
    theme: minty
    highlight: tango
    code-block-bg: true
    code-block-border-left: "#5BB867"
---

```{r config, include = FALSE}
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(napistu.r))

NAPISTU_R_CONFIG_NAME <- "napistu_r_config.yaml"
PROJECT_DIR <- "~/Desktop/DATA/Forny2023"
PPR_RESULTS_OUTFILE <- "ppr_results.tsv"
NETWORK_PLOTS_OUTDIR_NAME <- "network_plots"
OVERWRITE <- FALSE

checkmate::assert_file_exists(NAPISTU_R_CONFIG_NAME)
checkmate::assert_directory_exists(PROJECT_DIR)
checkmate::assert_string(PPR_RESULTS_OUTFILE)
checkmate::assert_string(NETWORK_PLOTS_OUTDIR_NAME)
checkmate::assert_logical(OVERWRITE, len = 1)

PPR_RESULTS_PATH <- file.path(PROJECT_DIR, PPR_RESULTS_OUTFILE)
# use the current directory and add plots to gitignore
# that way the pngs generated can be referenced and embedded in the rendered html
NETWORK_PLOTS_OUTDIR <- NETWORK_PLOTS_OUTDIR_NAME

checkmate::assert_file_exists(PPR_RESULTS_PATH)
checkmate::assert_directory_exists(NETWORK_PLOTS_OUTDIR_NAME)

# settings
MAX_SUBGRAPH_COMPONENTS <- 2L
```

Using Napistu from R involves creating a configuration which is used to set the Python environment used by reticulate and to load files produced by the Napistu python library. The assets that are needed for network analysis in R are:

1. An `SBML_dfs` object which serves as a relational database representation of the pathway representation.
2. A `NapistuGraph` object which is an `igraph.Graph` subclass formed by translating the `SBML_dfs` using a specified wiring strategy and attributes like edge weights.
3. A species identifiers table which contains the systematic identifiers for the molecular species in the model.
4. Optionally, precomputed distances between pairs of molecular species. This can greatly speed up calculations of shortest paths which is essential when calculating many such paths. In this notebook, we'll use the precomputed distances to calculate the molecular neighborhood around MMUT.

```{python, eval = FALSE}
from napistu.gcs import downloads

_ = downloads.load_public_napistu_asset(
    asset = "human_consensus_w_distances",
    data_dir = NAPISTU_DATA_DIR,
    subasset = "sbml_dfs"
)
```

```{r setup_napistu_py}
# load the sbml_dfs, network and other assets and bind a reticulate environment
# with napistu-py
napistu_config <- napistu.r::load_napistu_config(NAPISTU_R_CONFIG_NAME)
setup_napistu_list(napistu_config, skip_validation = TRUE)

# reticulate binding for the Napistu Python package
napistu <- napistu_list$python_modules$napistu

# `setup_napistu_list` creates a napistu_list object full of goodies
napistu_list
```

```{r refine_inputs}
# drop some uninformative pathways from the source counts table so they aren't highlighted in the network graph
# this could be handled upstream, but its easy enough to patch it here
napistu_list$reactions_source_total_counts <- napistu_list$reactions_source_total_counts %>%
  dplyr::filter(!(pathway_id %in% c("uncompartmentalization")), !stringr::str_detect(pathway_id, "/cache/"))
```

```{r load_ppr}
# load PPR scores and significance based on vertex permutation
ppr_results <- readr::read_csv(PPR_RESULTS_PATH, show_col_types = FALSE)
```

## Defining the neighborhood of MMUT

(R)-methylmalonyl-CoA CoA-carbonylmutase (MMUT) (an [isomerase in the TCA cycle](https://www.genome.jp/entry/R00833)) mutations are the root cause of a large fraction of MMA cases. To get warmed up on network visualization we can create a neighborhood around MMUT encompassing both MMUT's regulators and its downstream targets. Since the `NapistuGraph` we'll be querying includes a detailed human metabolic network, Recon3D, we can expect that the neighborhood will encompass both proteins and metabolites.

```{r yfg_selection}
# choose your favorite gene
YFG_ENSG <- "ENSG00000146085"

matching_ids <- napistu$ontologies$id_tables$filter_id_table(
  id_table = napistu_list$species_identifiers,
  identifiers = YFG_ENSG,
  bqbs = c("BQB_IS", "BQB_IS_ENCODED_BY")
)

S_ID <- purrr::pluck(matching_ids, "s_id")
```

```{r create_neighborhood_plots}
neighborhood_settings <- tibble::tribble(
  ~ alias, ~ max_steps, ~ max_neighbors, ~ network_layout, ~ out_width, ~ out_height,
  "big_picture", 5L, 200L, "kk", 20, 20,
  "close_up", 4L, 40L, "fr", 10, 10
) %>%
  dplyr::mutate(
    png_name = glue::glue("{YFG_ENSG}_{alias}_neighborhood.png"),
    png_path = file.path(NETWORK_PLOTS_OUTDIR, png_name),
    outfile_exists = file.exists(png_path) 
  )

if (OVERWRITE) {
  neighborhoods_to_plot <- neighborhood_settings
} else {
  neighborhoods_to_plot <- neighborhood_settings %>%
    dplyr::filter(!outfile_exists)
}

for (i in seq_len(nrow(neighborhoods_to_plot))) {
  
  settings <- neighborhoods_to_plot %>%
    dplyr::slice(i) %>%
    as.list()
  
  cli::cli_alert_info("Creating a neihborhood around {.field {YFG_ENSG}/{S_ID}} using {.field {settings$alias}} settings")
  
  neighborhoods <- napistu.r::create_neighborhood_table(
    napistu_list,
    species_id = S_ID,
    network_type = "hourglass",
    max_steps = settings$max_steps,
    max_neighbors = settings$max_neighbors
  )
  
  neighborhood_grob <- napistu.r::plot_neighborhoods(
    napistu_list,
    neighborhoods,
    # ignore weights until https://github.com/napistu/napistu/issues/15 is addressed
    edge_weights = NA,
    network_layout = settings$network_layout
  )
  
  ggplot2::ggsave(plot = neighborhood_grob, filename = settings$png_path, width = settings$out_width, height = settings$out_height)
}
```

```{r results='asis'}
generate_neighborhood_plot_blocks <- function(neighborhood_settings) {
  for (i in 1:nrow(neighborhood_settings)) {
    row <- neighborhood_settings[i, ]
    
    # Section header
    cat("### ", tools::toTitleCase(gsub("_", " ", row$alias)), "\n\n")
    
    # Display image
    cat('<img src="', row$png_path, '" style="width: 100%; display: block; margin: auto;" />\n\n', sep = "")
  }
}

generate_neighborhood_plot_blocks(neighborhood_settings)
```

## Defining subgraphs based on multiomic signatures

```{r select_enrichments}
TRAITS <- c(
  "OHCblPlus",
  "MMA_urine"
)
MEASURES <- c("log10_p", "stat")
MODALITIES <- c("transcriptomics", "proteomics")

attributes_of_interest <- tidyr::expand_grid(trait = TRAITS, measure = MEASURES, modality = MODALITIES) %>%
  dplyr::mutate(attr_name = glue::glue("{measure}_{trait}_{modality}"))

ppr_enriched_vertices <- ppr_results %>%
  dplyr::filter(
    q_value < 0.1,
    is_enriched,
    attribute %in% attributes_of_interest$attr_name
  ) %>%
  tidyr::nest(attr_scores = -attribute)
```

```{r create_subgraph_plots}
ppr_subgraph_settings <- ppr_enriched_vertices %>%
  dplyr::mutate(
    # plot the two largest weakly connected components of the induced subgraph
    # formed from vertices with net propagation enrichment
    png_name_1 = glue::glue("{attribute}_component_1.png"),
    png_path_1 = file.path(NETWORK_PLOTS_OUTDIR, png_name_1),
    # only check for existence of component 1 because there may only be 1 component
    outfile_exists = file.exists(png_path_1)
  )

if (OVERWRITE) {
  subgraphs_to_plot <- ppr_subgraph_settings
} else {
  subgraphs_to_plot <- ppr_subgraph_settings %>%
    dplyr::filter(!outfile_exists)
}

for (i in seq_len(nrow(subgraphs_to_plot))) {
  
  settings <- subgraphs_to_plot %>%
    dplyr::slice(i) %>%
    as.list()
  
  cli::cli_alert_info("Creating subgraph plots based on vertices enriched for {.field {settings$attribute}} signal")

  scores <- settings$attr_scores[[1]]
  
  # create an induced subgraph and select the largest weakly connected component
  
  subgraph <- napistu.r::define_subgraphs(
    napistu_list,
    subgraph_vertices = scores$vertex_name,
    max_components = MAX_SUBGRAPH_COMPONENTS
  )
  
  for (i in seq_along(subgraph)) {
    
    n_vertices <- length(subgraph[[i]]$component_graph$vs$get_attribute_values("name"))
    if (i != 1 && n_vertices < 50) {
      next
    }
    
    vertex_size <- 1/log(n_vertices, base = 4) * 2
    edge_width <- 1/log(n_vertices, base = 4) / 4
    # TO DO - try dropping edges outright once R package is reinstalled
    # or hide edges below a minimum weight value
    
    subgraph_grob <- napistu.r::plot_one_component(
      napistu_list,
      subgraph[[i]],
      network_layout = "kk",
      edge_weights = NA,
      vertex_size = vertex_size,
      edge_width = edge_width
    )
    
    outfile <- stringr::str_replace(settings$png_path_1, "_1.png", glue::glue("_{i}.png"))
    
    if (n_vertices > 1000) {
      out_width <- 20
      out_height <- 20
    } else {
      out_width <- 10
      out_height <- 10
    }
    
    ggplot2::ggsave(
      plot = subgraph_grob,
      filename = outfile,
      width = out_width,
      height = out_height
    )
  }
}
```

```{r results='asis'}
generate_subgraph_plot_blocks <- function(ppr_subgraph_settings) {
  for (i in 1:nrow(ppr_subgraph_settings)) {
    row <- ppr_subgraph_settings[i, ]
    
    # Section header
    cat("### ", tools::toTitleCase(gsub("_", " ", row$attribute)), "\n\n")
    
    # Display image
    cat('<img src="', row$png_path_1, '" style="width: 100%; display: block; margin: auto;" />\n\n', sep = "")
  }
}

generate_subgraph_plot_blocks(ppr_subgraph_settings)
```
